An SMT-based Approach to\\ Coverability Analysis

### 2014-03-17 paper/workspace/main.tex
Added submitted version of paper

\ifthenelse{\equal{\isDraft}{false}}{
  \title{An SMT-based Approach to Coverability Analysis}
  \author{
    Javier~Esparza\inst{1}
    \and
    Rupak~Majumdar\inst{2}
    \and
    Philipp~Meyer\inst{1}
    \and
    Filip~Niksic\inst{2}
    \and
    Rusl\'{a}n~Ledesma-Garza\inst{1}
  }
  \institute{
    Technische Universit\"at M\"unchen
    \and
    MPI-SWS,~Kaiserslautern~and~Saarbr\"ucken
  }

@@

  \begin{abstract}
    Model checkers based on Petri net coverability have been used
    successfully in recent years to verify safety properties of concurrent
    shared-memory or asynchronous message-passing software. The bottleneck
    in the verification process is the iterative computation of the
    (backward) reachable or (forward) coverable set, and most recent
    research has focused on non-trivial heuristics to speed it up.

    In this paper, we revisit a constraint approach to the problem based on
    classical Petri net analysis techniques: marking equation, place
    invariants, and traps. Given a marking M to be covered, the approach
    constructs constraints corresponding to increasingly precise
    overapproximations of the reachable markings covering M. If one of
    these constraints is unsatisfiable, then M is not coverable.

    Previous work on this approach suffered from the absence of efficient
    decision procedures for linear arithmetic, which nowadays exist as a
    part of SMT solvers. We show how to utilize an SMT solver to implement
    the constraint approach, and additionally, to generate a minimized
    inductive invariant from a safety proof.

    We empirically evaluate our procedure on a large set of existing Petri
    net benchmarks arising out of software verification. Surprisingly, even
    though our technique is incomplete, we show that it can quickly
    discharge all safe instances, usually competitive with, and occasionally
    much faster than, state-of-the-art techniques. Additionally, the
    inductive invariants computed are usually orders of magnitude smaller
    than those produced by existing solvers.
  \end{abstract}
}{
}

\input{sec_introduction}
\input{sec_illustration}
\input{sec_preliminaries}
\newpage
\input{sec_constraint_approach}
\newpage
\input{sec_invariants}
\newpage
\input{sec_experiments}
\input{sec_rel_work}
\input{sec_conclusions}

@@

@@

>>>>

\ifthenelse{\equal{\isDraft}{false}}{
  \title{An SMT-based Approach to\\
         Coverability Analysis}
  \author{
    Javier~Esparza\inst{1}
    \and
    Rusl\'{a}n~Ledesma-Garza\inst{1}
    \and
    Rupak~Majumdar\inst{2}
    \and
    Philipp~Meyer\inst{1}
    \and
    Filip~Niksic\inst{2}
  }
  \institute{
    Technische Universit\"at M\"unchen
    \and
    Max Planck Institute for Software Systems (MPI-SWS)
  }

@@

  \begin{abstract}
    Model checkers based on Petri net coverability have been used
    successfully in recent years to verify safety properties of concurrent
    shared-memory or asynchronous message-passing software. We revisit 
    a constraint approach to the problem based on
    classical Petri net analysis techniques. We show how to 
    utilize an SMT solver to implement
    the constraint approach, and additionally, to generate a minimized
    inductive invariant from a safety proof. We empirically evaluate 
    our procedure on a large set of existing Petri
    net benchmarks. Even though our technique is incomplete, 
    it can quickly discharge all safe instances. Additionally, the
    inductive invariants computed are usually orders of magnitude smaller
    than those produced by existing solvers.
  \end{abstract}
}{
}

\input{sec_introduction}
\input{sec_rel_work}
\input{sec_preliminaries}
\input{sec_safety}
\input{sec_safety_by_refinement}
%\input{sec_invariant}
\input{sec_invariant_by_refinement}
\input{sec_invariant_w_minimization}
\input{sec_experiments}
% \input{sec_conclusions}

@@

\newpage
\appendix
\input{appendix}

@@

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2014-03-17 paper/workspace/sec_experiments.tex
Added submitted version of paper

\section{Experiments}

To evaluate the described methods, we implemented them in a tool called
\emph{Petrinizer}. Petrinizer is essentially a set of Bash and Prolog scripts
that mediate between the input, specified in the input format of
MIST2\footnote{\url{https://github.com/pierreganty/mist}}, and the SMT
solver Z3\footnote{\url{http://z3.codeplex.com/}} \cite{DeMouraTACAS08}. With
the choice of Bash and Prolog over languages with bindings for Z3 API, we
restricted Z3's full power. Indeed, by using Z3 as an external tool, we
were not able to fully exploit the incremental nature of its linear arithmetic
solver (TODO: cite appropriate papers). However, even the non-optimal
implementation turned out to be robust and scalable, with the main advantage of
being able to tackle problem instances that are out of reach for
state-of-the-art coverability checkers.

The evaluation of Petrinizer had three main goals. First, we wanted to compare
its performance against state-of-the-art tools like MIST2, BFC
and IIC (cite!). Second, as the method it
implements is incomplete, we wanted to measure its rate of success on safe problem instances. As a
subgoal, we wanted to investigate the usefulness and necessity of traps, as
well as to what extent real arithmetic suffices over integer arithmetic in
proving safety. And last, since the language of linear arithmetic, used by
Petrinizer, allows for a more succint representation of formulas than the
language used by IIC, we wanted to compare sizes of invariants the two tools produce.

Inputs that were used in tests come from various sources (TODO: write a
sentence to justify the variety).
One source is the collection of Petri nets from the literature that is part of the MIST2 toolkit.
The second source are Petri nets originating from the analysis of concurrent C
programs that were used in the evaluation of BFC \cite{KaiserCONCUR12}. Then
there are inputs originating from the provenance analysis of a medical and a
bug-tracking system \cite{MajumdarSAS13}. Finally, we generated a set of Petri
nets from Erlang programs, using an Erlang verification tool Soter
\cite{DOsualdoSAS13}. (TODO: Acknowledge Emanuele D'Osualdo for help.)

All experiments were performed on the identical machines, equipped with
Intel Xeon 2.66 GHz CPUs and 48 GB of memory, running Linux 3.2.48.1 in 64-bit
mode. Execution time was limited to 100,000 seconds (27 hours, 46 minutes and
40 seconds), and memory to 2 GB.

\subsection{Performance}

Explain what we observe:
\begin{itemize}
  \item On small examples IIC mostly outperforms Petrinizer. Probably due
    to a slower parser, setting up and invoking Z3.
  \item On examples of a similar size, coming from the same source,
    unlike with IIC, resource consumption is fairly uniform. (Medical examples.)
  \item Petrinizer is able to handle some huge examples that are out of
    scope for IIC.
  \item There were no cases when IIC was able to finish, but Petrinizer ran
    out of resources. As a matter of fact, Petrinizer finishes in all cases,
    whereas other tools somethimes fail.
\end{itemize}

\emph{Invariant sizes.} Again, explain how we measure it. For IIC, we measure
the number of non-null literals, the number of clauses, total length of
the invariant in characters. For Petrinizer, we measure the number of
non-null coefficients in the linear expressions and the total length of the
invariant in characters. Argue why it makes sense to compare any of these
quantities. Point out the difference in size with and without the invariant
minimization.

\emph{Rate of success.} Point out that Petrinizer mostly successfully proves
safety. Specifically, it proves safety in all of the safe Soter examples.
Point out the cases where traps were actually useful, and potentially progress
into some pseudo-philosophical discussion on what constitutes a good set of examples.

>>>>

\section{Experimental Evaluation}
\label{sec_experiments}

We have implemented our algorithms in a tool called
\emph{Pe\-tri\-ni\-zer}. 
Petrinizer is implemented as a combination of Bash and Prolog scripts,
and uses Z3 \cite{MouraB08} as a backend SMT solver.
It takes as input coverability problem instances encoded in the MIST
input format\footnote{\url{https://github.com/pierreganty/mist}}, and it
runs one of the selected methods. 
We implemented all possible combinations of
methods: with and without trap refinement, with real and integer arithmetic,
with and without invariant construction, with and without invariant
minimization.

The evaluation of Petrinizer had two main goals. First, as the underlying
methods are incomplete, we wanted to measure their success rate on standard
benchmark sets. 
As a subgoal, we wanted to investigate
the usefulness and necessity of traps, the benefit of using integer arithmetic
over real arithmetic, and the sizes of the constructed invariants.
The second goal was to measure Petrinizer's performance and to compare it
with state-of-the-art tools: IIC \cite{KloosMNP13}, BFC\footnote{The most recent version of BFC at the time of
writing the paper was 2.0. However, we noticed it sometimes reports
inconsistent results, so we used version 1.0 instead. The tool can be
obtained at \url{http://www.cprover.org/bfc/}.} \cite{KaiserKWCONCUR12}, and
MIST\footnote{MIST consists of several methods, most of them based on
  EEC \cite{GeeraertsRB06}. We used the abstraction
  refinement method that tries to minimize the number of places in the net \cite{GantyFI08}.}.

\paragraph{Benchmarks.} For the inputs used in the experiments, we collected a
large number of coverability problem instances originating from various sources. The
collection contains 178 examples, out of which 115 are safe, and is
organized into five example suites. The first suite is a collection of
Petri net examples from the MIST toolkit. This suite contains a
mixture of 29 examples, both safe and unsafe. It contains both real-world and
artificially created examples. The second suite consists of 46 Petri nets
that were used in the evaluation of BFC \cite{KaiserKWCONCUR12}. They
originate from the analysis of concurrent C programs, and they are mostly
unsafe. The third and the forth suite come from the
provenance analysis of messages in a medical and a bug-tracking system
\cite{MajumdarSAS13}. The medical suite contains 12 safe
examples, and the bug-tracking suite contains 41 examples, all
safe except for one. The final, fifth suite contains 50 examples that
come from the analysis of Erlang programs \cite{DOsualdoSAS13}. We generated
them ourselves using an Erlang verification tool Soter \cite{DOsualdoSAS13},
from the example programs found on Soter's
website\footnote{http://mjolnir.cs.ox.ac.uk/soter/}. Out of 50 examples in this
suite, 38 are safe. This suite also contains the largest example in the
collection, with 66,950 places and 213,635 transitions.
For our evaluation, only the 115 safe instances are interesting.

\begin{table}[t]
  \centering
  \caption{Safe examples that were successfully proved safe.}
  \label{table:rate-of-success}
  \scriptsize
  \begin{tabular}{|l|r|r|r|r|r|r|r|}
    \hline
    \multicolumn{1}{|c|}{Suite} & \multicolumn{1}{|c|}{Safety}
    & \multicolumn{1}{|c|}{Ref./reals} & \multicolumn{1}{|c|}{Ref./ints}
    & \multicolumn{1}{|c|}{IIC} & \multicolumn{1}{|c|}{BFC}
    & \multicolumn{1}{|c|}{MIST} & \multicolumn{1}{|c|}{Total} \\
    \hline
    MIST        & 14 & 20 & 20 & {\bf 23} & 21 & 19 & 23 \\
    BFC          & {\bf 2} & {\bf 2} & {\bf 2} & {\bf 2} & {\bf 2} & {\bf 2} & 2 \\
    Medical      & 4 & 4 & 4 & 9 & {\bf 12} & 10 & 12 \\
    Bug-tracking & {\bf 32} & {\bf 32} & {\bf 32} & 0 & - & 0 & 40 \\
    Erlang       & 32 & 36 & {\bf 38} & 17 & 26 & 2 & 38 \\
    \hline
    Total        & 84 & 94 & {\bf 96} & 51 & 61 & 33 & 115 \\
    \hline
  \end{tabular}
\end{table}

\paragraph{Rate of success on safe examples.} As shown in Table
\ref{table:rate-of-success}, even with the weakest of the methods---safety based
on marking equation over reals---Petrinizer is able to prove safety for 84 out
of 115 examples. Switching to integer arithmetic does not help: The number of
examples proved safe remains 84. Using refinement via traps, Petrinizer proves
safety for 94 examples. Switching to integer arithmetic in this case helps:
Another two examples are proved safe, totaling with 96 out of 115 examples. In
contrast to these numbers, the most successful existing tool turned out to be BFC, proving safety for
only 61 examples\footnote{This number comes with a caveat: We were not able to
run BFC on the bug-tracking suite, as the examples were too big for the tool that translates
MIST format into BFC format. For the BFC and Erlang suites, which contain
some similarly large or even larger examples, we initially had the inputs in BFC
format, and translated them into MIST format, thus avoiding the issue. Translation
utilities are provided by the authors of BFC.}. The small numbers are due to
time and memory limits. Even though the methods these tools implement
are complete, the tools themselves are limited by the time and space they can
use.

Looking at the results accross different suites, we see that Petrinizer
performed poorest on the medical suite, proving safety for only 4 out of 12
examples. On the other hand, on the bug-tracking suite, which was completely
intractable for other tools, it proved safety for 32 out of 40 examples.
Furthermore, using traps and integer arithmetic, Petrinizer successfuly proved
safety for all safe Erlang examples. We find this result particularly
surprising, as the original verification problems for these examples seem non-trivial.

\begin{figure}[t]
  \centering
  \input{fig_invariant_size}
  \caption{Graph on the left shows a relation of sizes of constructed
  invariants to the number of places in the corresponding Petri nets. Graph
  on the right shows comparison in size of invariants produced by Petrinizer and
  IIC. Axes represent size on a logarithmic scale. Each dot represents one example.}
  \label{fig:invariant-size}
\end{figure}

\paragraph{Invariant sizes.} As the invariant construction is a new addition to
the existing methods, we wanted to inspect sizes of the constructed invariants.
We took the number of atomic terms appearing in an invariant's linear expressions
as a measure of its size. When we relate sizes of invariants to number of places
in the corresponding Petri net (left graph in Fig.~\ref{fig:invariant-size}), we
see that invariants are most of the times very succinct. As an example, the
largest invariant had 814 atomic terms, and the corresponding Petri net, coming
from the Erlang suite, had 4,763 places. For the largest Petri net in the
collection, the one with 66,950 places, the constructed invariant had 339 atomic
terms.

We also compared sizes of constructed invariants with
sizes of invariants produced by IIC \cite{KloosMNP13}. IIC's invariants
are expressed as CNF formulas over atoms of the form $x<a$,
for a variable $x$ and a constant $a$. As a measure of size for these formulas,
we took the number of atoms they contain. As the right graph in
Fig.~\ref{fig:invariant-size} shows, when compared to IIC's invariants, Petrinizer's invariants
are never larger, and are often orders of magnitude smaller.

\paragraph{Performance.} To ensure accuracy and fairness, all experiments were
performed on identical machines, equipped with Intel Xeon 2.66 GHz CPUs and 48 GB of memory, running Linux 3.2.48.1
in 64-bit mode. Execution time was limited to 100,000 seconds (27
hours, 46 minutes and 40 seconds), and memory to 2 GB.

Due to dissimilarities between the compared tools, selecting a fair
measure of time turned out to be tricky. On the one hand, as
Petrinizer communicates with Z3 via temporary files, it spends a
considerable amount of time doing I/O operations.  On the other hand,
as BFC performs both a forward and a backward search, it naturally
splits the work into two threads, and runs them in parallel on two CPU
cores. In both cases, the actual elapsed time does not quite
correspond to the amount of computational effort we wanted to measure.
Therefore, for the measure of time we selected the \emph{user time},
as reported by the \emph{time} utility on Linux. User time measures
the total CPU time spent executing the process and its children. In
the case of Petrinizer, it excludes the I/O overhead, and in the case
of BFC, it includes total CPU time spent on both CPU cores.

\begin{figure}[t]
  \centering
  \input{fig_performance_various}
  \caption{Time overhead of trap refinement, invariant
  construction and invariant minimization. Axes represent time in seconds on a
  logarithmic scale. Each dot represents
  execution time on one example. The graph on the left
  only shows examples for which at least one trap appeared in the refinement.
  Similarly, the other two graphs only show safe examples.}
  \label{fig:performance-various}
\end{figure}

\paragraph{Time overhead of Petrinizer's methods.} Before comparing Petrinizer
with other tools, we analyze time overhead of trap refinement, invariant construction, and invariant minimization.
Fig.~\ref{fig:performance-various} summarizes the results.
The leftmost graph in Fig.~\ref{fig:performance-various} shows that traps
incur a significant overhead. This is not too surprising as, each time
a trap is found, the main system has to be updated with a new trap
constraint and solved again. Thus the actual overhead depends on the number
of traps that appear during the refinement. In the experiments, there
were 32 examples for refinement with integer arithmetic where traps
appeared at least once. The maximal number of traps in a single example was 9.

In the case of invariant construction, as shown on the middle graph in
Fig.~\ref{fig:performance-various}, the overhead is more
uniform and predictable. The reason is that constructing the invariant
involves solving the dual form of the main system as many times as
there are disjuncts in the property violation constraint. In most
cases, the property violation constraint has one disjunt. A single example
with many disjuncts, having 8989 of them, appears on the graph
as an outlier.

In the case of invariant minimization, as the rightmost
graph in Fig.~\ref{fig:performance-various} shows, time overhead is
quite severe.
The underlying data contains examples of slowdowns of up to 30$\times$.
At the same time, the added benefit of minimization is negligible: There were only four examples where the invariant
was reduced, and that was for mere 2-3\%. 
Therefore, invariant minimization does not seem to pay off for these examples.

\begin{figure}[t]
  \centering \input{fig_performance_vs_tools}
  \caption{Comparison of execution time for Petrinizer vs.~IIC,
  BFC and MIST. Graphs in the top row show comparison in the case without
  invariant construction, and graphs in the bottom row show comparison in
  the case with invariant construction. Axes represent time in seconds on a
  logarithmic scale. Each dot represents execution time on one example.}
  \label{fig:performance-vs-tools}
\end{figure}

\paragraph{Comparison with other tools.} Six graphs in
Fig.~\ref{fig:performance-vs-tools} show the comparison of execution time for Petrinizer vs.~IIC, BFC, and MIST. 
In the comparison, we used
the refinement methods, both with and without invariant construction.
In general, we observe that other tools outperform
Petrinizer on small examples, an effect that can be explained by the
overhead of starting Bash and Prolog interpreters and Z3. 
However, on large examples Petrinizer consistently outperforms other tools.
Not only does it finish in all cases within the given time and memory
constraints, it even finishes in under 100
seconds in all but two cases. The two cases are the
large example from the Erlang suite, with 66,950 places and 213,635
transitions and, in the case of invariant construction, the example
from the MIST suite, with 8989 disjuncts in the property violation constraint.

\paragraph{Conclusions.}
%
Marking equations and traps are classical techniques in Petri net theory, but have fallen out of favor
in recent times in comparison with state-space traversal techniques in combination with abstractions
or symbolic representations.
Our experiments demonstrate that, when combined with the power of a modern SMT solver, these techniques
can be surprisingly effective in finding proofs of correctness (inductive invariants) of common benchmark examples arising
out of software verification.
It remains to be seen whether these techniques can be used to check liveness properties of concurrent
software, which often reduce to Petri net \emph{reachability} problems and are out of scope of other
state-space traversal techniques.

\paragraph{Acknowledgements.}
%
We thank Emanuele D'Osualdo for help with the Soter tool.

### 2014-03-17 paper/workspace/sec_introduction.tex
Added submitted version of paper

This paper makes the following contributions.
\begin{itemize}
\item An application of \zthree\ to the constraint approach and its
  corresponding evaluation.
\item A method that given a network and a property constructs a
  minimal invariant.
\end{itemize}

>>>>

In recent years many papers have proposed and developed techniques for
the verification of concurrent software
\cite{KaiserKW10,GantyM12,BouajjaniEmmi12,KaiserKWCONCUR12,DOsualdoSAS13}.
In particular, model checkers based on Petri net coverability have been successfully
applied.  Petri nets are a natural model for this problem for two
resons.  The first and obvious one is that they are a simple and
natural automata-like model for concurrent systems. The second, more
specific, is that they can model certain programs with an unbounded
number of threads or thread creation. In a nutshell, the places of the
net correspond to program points, and the number of tokens in a place
models the number of threads that are currently at that point. This
point, which was first observed in \cite{GS92}, was later revisited in
\cite{DelzannoRB02}, and, more implicitly, in \cite{KaiserKW10}.

The problem whether at least one thread can reach a given program
point (modelling some kind of error), naturally reduces to the {\em
  coverability problem} for Petri nets: given a net $N$ and a marking
$M$, decide whether some reachable marking of $N$ {\em covers} $M$,
i.e., puts at least as many tokens as $M$ on each place. While the
decidability and EXPSPACE-completeness of the coverability problem
were settled long ago \cite{KarpM69,Rackoff78}, new algorithmic ideas
have been developed in recent years
\cite{GeeraertsRB06,GantyFI08,ValmariH12,KaiserKWCONCUR12,KloosMNP13}. 
The techniques are based on forward or backward state-space exploration, 
which is accelerated in a number of ways in order to cope with the possibly 
infinite number of states.

In this paper we revisit an approach to the coverability problem based
on classical Petri net analysis techniques: the marking equation and
traps \cite{murata1989petri,EsparzaM00,Reisig13}.
The marking equation is a system of linear constraints that can be
easily derived from the net, and whose set of solutions
overapproximates the set of reachable markings.  The marking equation
can be supplemented with linear constraints specifying a set of unsafe
markings, and solved using standard linear or integer programming. If
the equations are infeasible, then all reachable markings are safe. If
not, then, as suggested in \cite{EsparzaM00}, one can use a
counterexample to derive an additional constraint in the shape of a
{\em trap}: a set of places that, loosely speaking, once marked cannot
be ``emptied''.

While these techniques are well known, their application has always
suffered from the absence of efficient decision procedures for 
linear arithmetic together with Boolean satisfiability.  
Moreover, and to the best of our knowledge, their
systematic application to verification of concurrent software has not
been studied yet. Profiting from recent advances in SMT-solving
technology, we reimplement the techniques on top of the Z3 SMT solver
\cite{MouraB08}, and apply them to a large collection of benchmarks.

Our first and surprising finding is that, while the technique is theoretically not
complete, it is powerful enough to prove safety of 96 out of a total
of 115 safe benchmarks gathered from current research papers in concurrent software
verification. In contrast, state-of-the-art tools for coverability
proved only 61, 51, or 33 of these 115 cases!
Moreover, and possibly due to the characteristics of the
application domain, even the simplest version of the technique ---based on the marking equation--- is
successful in 84 cases. 
As a second contribution, and inspired
by work on interpolation, we show that a dual version of the classical
set of constraints, equivalent in expressive power, can be used not
only to check safety, but to produce an inductive invariant.
While some existing solvers based on state-space
exploration can also produce such invariants, we show that inductive invariants obtained
through our technique are usually orders
of magnitude smaller.  Additionally, while we can use the 
SMT solver iteratively to minimize the invariant, the tool almost always
provides a minimal one at the first attempt.

% The paper is organized as follows.
% We introduce the basic notation in Section~\ref{sec_preliminaries}.
% We recall the marking equation and traps approach in
% Sections~\ref{sec_method_safety}
% and~\ref{sec_method_safety_by_refinement}.
% We introduce three novel methods to constructing invariants in
% Sections~\ref{sec_method_invariant}
% % \ref{sec_method_invariant_by_refinement},
% and
% \ref{sec_method_invariant_w_minimization}.
% Section~\ref{sec_experiments} presents (1) our experimental evaluation of
% the application of SMT solver \zthree\ to the constraint-based
% approach, and (2) our experimental evaluation of our methods for
% invariant construction.
% Section~\ref{sec_rel_work} reviews related
% work. Section~\ref{sec_conclusions} presents our conclusions.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2014-03-17 paper/workspace/sec_preliminaries.tex
Added submitted version of paper

\section{Preliminaries}

\begin{verbatim}
* Petri net: definition
 N = (S, T, E, M0), where
     S : places
     T : transitions
     E : edges
     M0: initial marking

* Marking: definition

* Visual representation of Petri Nets: description, example

* Property P: safety property expressed over linear arithmetic formulas.

* Linear arithmetic formula: examples

* N \models P: definition

* Whatever you believe should be defined for the paper.
\end{verbatim}

\iffalse
\begin{verbatim}
 Examples:
  x + y < 0
  x + y < 0 \\and x + z > 0
  x + y < 0 \\or  x + z > 0

 ~P = (x + y <  0 \\and x + z >  0) \\or  x + z >  0
  P = (x + y >= 0 \\or  x + z <= 0) \\and x + z <= 0
\end{verbatim}
\fi

>>>>

\section{Preliminaries}
\label{sec_preliminaries}

A \emph{Petri net} is a tuple $(P, T, F, m_0)$, where $P$ is a set of
\emph{places}, $T$ is a (disjoint) set of \emph{transitions},
$F : (P\times T) \cup (T\times P) \to \set{0,1}$ is the \emph{flow function},
and $m_0: P \rightarrow \mathbb{N}$ is the initial marking.
%
For $x\in P\cup T$, the \emph{pre-set} is
$\pre{x}=\{y\in P\cup T\mid F(y,x) =1\}$
and the \emph{post-set} is $\post{x}=\{y\in P\cup T\mid F(x,y) = 1\}$.
We extend the pre- and post-set to a subset of $P \cup T$ as the union of
the pre- and post-sets of its elements.

A \emph{marking} of a Petri net is a function $m\colon P \to \mathbb{N}$,
which describes the number of tokens $m(p)$ in 
each place $p\in P$. 
Assuming an enumeration $p_1, \ldots, p_n$ of $P$, we
often identify $m$ and the vector $(m(p_1), \ldots, m(p_n))$.
For a subset $P'\subseteq P$ of places, we write $m(P') = \sum_{p\in P'} m(p)$.

A transition $t \in T$ is \emph{enabled at $m$} iff
for all $p \in \pre{t}$, we have $m(p) \ge F(p, t)$.
A transition $t$ enabled at $m$ may \emph{fire},
yielding a new marking $m'$ (denoted $m \xrightarrow{t} m'$),
where $m'(p) = m(p) + F(t,p) - F(p,t)$.

A sequence of transitions, $\sigma = t_1 t_2 \ldots t_r$ is an
$\emph{occurrence sequence}$ of $N$ iff there exist markings
$m_1, \ldots, m_r$ such that $m_0 \xrightarrow{t_1} m_1
\xrightarrow{t_2} m_2 \ldots \xrightarrow{t_r} m_r$. The marking
$m_r$ is said to be \emph{reachable} from $m_0$ by the occurrence
of $\sigma$ (denoted $m_0 \xrightarrow{\sigma} m_r$).

A \emph{property} $\varphi$ is a linear arithmetic constraint over the free variables $P$. 
The property $\varphi$ holds on a marking $m$ iff $m \models \varphi$.
% Examples of properties are $s_1 \le 2$, $s_1 + s_2 \ge 1$ and
% $((s_1 \le 1) \land (s_2 \ge 1)) \lor (s_3 \le 1)$.
% 
A Petri net $N$ satisfies a property $\varphi$ (denoted by $N \models \varphi$)
iff for all reachable markings $m_0 \xrightarrow{\sigma} m$, we have
$m \models \varphi$.
A property $\varphi$ is an \emph{invariant} of $N$ if it holds for every reachable marking.
A property is inductive if whenever $m\models \varphi$ and $m \xrightarrow{t} m'$ for some $t\in T$
and marking $m'$, we have $m'\models \varphi$.

% A trap is a set of places $S \subseteq P$ such that $\post{S} \subseteq \pre{S}$.
% If a trap $S$ is marked in $m_0$, i.e. $\sum_{p \in S} m_0(p) > 0$, then it is also marked in all reachable markings.

Petri nets are represented graphically as follows: places and transitions
are represented as circles and boxes, respectively. For $x, y \in P \cup T$,
there is an arc leading from $x$ to $y$ if{}f $F(x,y)=1$.

\input{fig_lamport_src}
\input{fig_lamport_net}

Petri nets can be used to model behavior of concurrent software. As an
example, consider Lamport's 1-bit
algorithm for mutual exclusion \cite{LamportJACM86}, shown in
Fig.~\ref{fig_lamport_src}. 
A Petri net model for the code is shown in Fig.~\ref{fig_lamport_net}.
The two grey blocks model the control flow of the two
processes. For instance, the token on place $p_1$ models the current
position of process 1 at program location $p_1$. The three places in the middle of the
diagram model the current values of the variables. For instance, a
token on place $notbit_1$ indicates that the variable ${\it bit}_1$ is currently
set to ${\it false}$.
The mutual exclusion property, which states that the two processes cannot be in the
critical section at the same time, corresponds to the property that
places $p_3$ and $q_5$ cannot both have a token at the same time.
